# Lesson 3

Today: 

(1) More graphing examples
(2) How R works

If your analysis is to be a convincing, the trail from data to final output must be open and available to all. Markdown helps you create scientific reports that are a mixture of text and code. This makes it easy to create an understandable trail from hypothesis, to data, to analysis, to results. Reproducible research.

## Scatter plots

Functions from the {ggplot2} package are used to make graphs. We make these graphing functions available for a given session of R (every time we open RStudio) with the `library(ggplot2)` function.

As an example, consider the `airquality` data frame. The data contains daily air quality measurements from a location in New York City between May and September of 1973.

Follow along by pressing the green arrows when we get to a code chunk.
```{r}
airquality
```

The data contains 153 rows and 6 columns. Each row is a set of measurements across six variables on a given day.

Most data we will work with are like this. Each row is a set of measurements (a case) and each column is a variable.

The columns (variables) include the measurements of ozone concentration (`Ozone`) (ppm), solar radiation (`Solar.R`) (langley), wind speed (`Wind`) (mph), temperature (`Temp`) (F), as well as `Month` and `Day`.

Question: Are ozone concentrations higher on warmer days? Let's see what the data say.

The scatter plot is one of the most useful statistical graphs. It describes the _relationship_ between two variables. It is made by plotting the variables in a plane defined by the values of the variables. 

Using the {ggplot2} functions, we answer the question above by mapping the `Temp` variable to the x aesthetic and the `Ozone` variable to the y aesthetic.

More simply we could say that we plot `Temp` on the x axis and `Ozone` on the y axis. Put we want to recognize that the axes are aesthetics (there are other aesthetics like color, size, etc).
```{r}
library(ggplot2)

ggplot(data = airquality) + 
  geom_point(mapping =  aes(x = Temp, y = Ozone))
```

What do you see? Why the warning?

To help us better describe the relationship we add another layer. This layer is defined by `geom_smooth()` which takes the same aesthetics.
```{r}
ggplot(data = airquality) + 
  geom_point(mapping =  aes(x = Temp, y = Ozone)) +
  geom_smooth(mapping =  aes(x = Temp, y = Ozone))
```

The smooth line describes how the average ozone concentration varies with temperature. For lower temperatures there is not much change in ozone concentrations as temperatures increase, but for higher temperatures the increase in ozone concentrations is more pronounced.

In the above code we used the same mapping for the point layer and the smooth layer. We can simplify the code by putting the `mapping =` argument into the `ggplot()` function.
```{r}
ggplot(data = airquality,
       mapping =  aes(x = Temp, y = Ozone)) + 
  geom_point() +
  geom_smooth()
```

Question: On average is ozone concentration higher on windy days? Create a graph to help you answer this question.
```{r}
ggplot(data = airquality, 
       mapping = aes(x = Wind, y = Ozone)) + 
  geom_point() +
  geom_smooth()
```

What is the answer?

We can use a label instead of a dot for the locations in this two-dimensional scatter plot by adding the `label` aesthetic and using `geom_text`. 
```{r}
ggplot(data = airquality, 
       mapping = aes(x = Wind, y = Ozone, label = Ozone)) +
  geom_text()
```

To map an aesthetic to a variable, associate the name of the aesthetic to the name of the variable inside `aes()`.

We can make the plot interactive by using the `ggplotly()` function from the {plotly} package. We simply put the above code inside this function.
```{r}
plotly::ggplotly(
  ggplot(data = airquality, 
         mapping =  aes(x = Temp, y = Ozone)) + 
  geom_point() +
  geom_smooth()
)
```

Hover/zoom etc.

As another example, consider the Palmer penguin data set from https://education.rstudio.com/blog/2020/07/palmerpenguins-cran/. 

The data are located on the web at the following URL. We first save the location as an object called `loc`.
```{r}
loc <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
```

Note that this object is now located in our environment. It is simply a string of characters (letters, backslashes, etc) in quotes. A character object.

Next we get the data and save it as an object called `penguins` with the `read.csv()` function. Inside the parentheses of the function we put the name of the location. 
```{r}
penguins <- read.csv(loc)
```

We note that the object `penguins` is now in our environment. It is a data frame containing 344 rows (observations) and 8 variables. We list the first 10 rows and 7 columns by typing the name of the object as follows.
```{r}
penguins
```

The data are 344 individual penguins each described by species (Adelie, Chinstrap, Gentoo), where it was found (island name), length of bill (mm), depth of bill (mm), body mass (g), male or female, and year.

Each penguin belongs to one of three species. To see how many of the 344 penguins are in each species we use the `table()` function. Between the parentheses of this function we put the name of the data `penguins` followed by the `$` sign followed by the name of the column `species`.
```{r}
table(penguins$species)
```

Said another way, we reference columns in the data with the `$` sign so that `penguins$species` is how we refer to the column `species` in the data object named `penguins`.

There are 152 Adelie, 68 Chinstrap, and 124 Gentoo penguins.

We plot the relationship between flipper length and body mass for each of the three species.
```{r}
ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()  
```

Penguin flipper length and body mass show a positive relationship (association). Penguins with longer flippers tend to be larger.

How does this positive relationship vary by species?

We answer this question with another aesthetic. We assign a level of the aesthetic (here a color) to each unique value of the variable, a process known as scaling. The `ggplot()` function also adds a legend that explains which levels correspond to which values.

```{r}
ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) 
```

Returning to the `mpg` data set from last time.
```{r}
ggplot(data = mpg,
       mapping = aes(x = displ, y = hwy, color = class)) + 
  geom_point()
```

The colors reveal that the unusual points (on the right side of the plot) are two-seaters. Sports cars have large engines like SUVs and pickup trucks, but small bodies like midsize and compact cars, which improves their gas mileage. In hindsight, these cars were unlikely to be hybrids since they have large engines.

In the above example, we mapped `class` to the color aesthetic, but we could have mapped `class` to the shape aesthetic, which controls point shapes.
```{r}
ggplot(data = mpg,
       mapping = aes(x = displ, y = hwy, shape = class)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

What happened to the SUVs? The `ggplot()` function will only use six shapes at a time. By default, additional groups will go un-plotted when you use the shape aesthetic.

For each aesthetic, you use `aes()` to associate the name of the aesthetic with a variable to display. The `aes()` function gathers together each of the aesthetic mappings used by a layer and passes them to the layer's mapping argument. 

The syntax highlights a useful insight about `x` and `y`: the x and y locations of a point are themselves aesthetics, visual properties that you can map to variables to display information about the data. 

We can also set the aesthetic properties of your geom manually. For example, we can make all of the points in our plot blue.
```{r}
ggplot(data = mpg,
       mapping = aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")
```

Here, the color doesn't convey information about a variable, but only changes the appearance of the plot. To set an aesthetic manually, set the aesthetic by name as an argument of your geom function; i.e. it goes outside of `aes()`. You'll need to pick a level that makes sense for that aesthetic:

* The name of a color as a character string (with quotes).
* The size of a point in millimeters.
* The shape of a point as a number, as shown below.

R has 25 shapes that are identified by numbers. There are some seeming duplicates: for example, 0, 15, and 22 are all squares. The difference comes from the interaction of the `color` and `fill` aesthetics. The hollow shapes (0--14) have a border determined by `color`; the solid shapes (15--18) are filled with `color`; the filled shapes (21--24) have a border of `color` and are filled with `fill`.

### Facets

Another way to add additional variables which is useful for categorical variables, is to split the plot into facets (subplots that each display one subset of the data). A categorical variable is one that can take only a limited, and usually fixed, number of possible values.

To facet your plot by a categorical variable, use `facet_wrap()`. The first argument of `facet_wrap()` is a formula, which you create with `~` followed by a variable name (here "formula" is the name of a data structure in R). The variable that you pass to `facet_wrap()` should only have a limited number of values (categorical). 

The variable `class` in the data frame `mpg` is a character string. You can see this by typing
```{r}
str(mpg)
```

There are seven car classes. We put `class` in the `facet_wrap()` function. Everything is the same as before on the first two code lines but we add the `facet_wrap()` function.
```{r}
ggplot(data = mpg,
       mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_wrap(~ class, nrow = 2) 
```

The plot shows separate scatter plots one for each of the seven classes. More on graphs later.

## Calculations

Let's see how we can do some arithmetic in R.

R evaluates commands typed at the prompt and returns the result to the screen. The prompt is the blue greater than symbol (`>`). To find the sum of the square root of 25 and 2, at the prompt type
```{r}
sqrt(25) + 2
```

The number inside the brackets indexes the output. Here there is only one bit of output, the answer 7. The prompt that follows indicates R is ready for another command.
```{r}
12/3 - 5
```

How would you calculate the 5th power of 2? How would you find the product of 10.3 & -2.9? How would you find the average of 8.3 and 10.2?

How about 4.5% of 12,000?
```{r}
.045 * 12000 
```

## Functions

Many math and statistical functions are available. A function has a name followed by a pair of parentheses. Arguments are placed inside the parentheses as needed.

For example,
```{r}
sqrt(2)
```

```{r}
sin(pi)
```

How do you interpret this output?  Type (highlight then click Run): .0000000000000001224647
Why not zero? What does the `e-16` mean?

```{r}
exp(1)
log(10)
```

Many functions have arguments with default values. For example, you only need to tell the random number generator `rnorm()` how many numbers to produce. The default mean is zero. To replace the default value, specify the corresponding argument name.
```{r}
rnorm(10)
rnorm(10, mean = 5)
```

## Syntax is important

You get an error message when you type a function that R does not understand. For example:
```{r eval=FALSE}
squareroot(2)
```
Error: could not find function "squareroot"

```{r eval=FALSE}
sqrt 2
```
Error: syntax error

```{r}
sqrt(-2)
```

```{r eval=FALSE}
sqrt(2
```

The last command shows what happens if R encounters a line that is not complete. The continuation prompt (`+`) is printed, indicating you did not finish the command.

## Saving an object

Use the assignment operator to save an object. You put a name on the left-hand side of the left pointing arrow (`<-`) and the value on the right. Assignments do not produce output.
```{r}
x <- 2 
x + 3    
x <- 10
```

Here you assigned `x` to be a numeric object. Assignments are made using the left-pointing arrow (less than followed by a dash) [or an equal sign.]

## Object names

You are free to make object names out of letters, numbers, and the dot or underline characters.  A name starts with a letter or a dot (a leading dot may not be followed by a number). But you can't use mathematical operators, such as +, -, *, and /.

Some examples of names include:
```{r}
x <- 2
n <- 25
a.long.number <- 123456789
ASmallNumber <- .001
```

Case matters. `DF` is different than `df` or `Df`.

Some names are commonly used to represent certain types of data. For instance, `n` is for length; `x` or `y` are data vectors; and `i` and `j` are integers and indices.

These conventions are not forced, but consistent use of them makes it easier for you (and others) to understand what you've done.

## Entering data

The `c()` function is useful for getting a small amount of data into R. The function combines (concatenates) items (elements). Example: consider a set of hypothetical annual land falling hurricane counts over a ten-year period.

2  3  0  3  1  0  0  1  2  1

To enter these into your environment, type
```{r}
counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)
counts
```

Notice a few things. You assigned the values to an object called counts. The assignment operator is an equal sign (`=`). Values do not print. They are assigned to an object name. They are printed by typing the object name as we did on the second line. Finally, the values when printed are prefaced with a `[1]`. This indicates that the object is a vector and the first entry in the vector is a value of 2 (The number immediately to the right of [1]). More on this later.

You can save some typing by using the arrow keys to retrieve previous commands.  Each command is stored in a history file and the up arrow key will move backwards through the history file and the down arrow forwards.  The left and right arrow keys will work as expected.
 
## Applying a function

Once the data are stored in an object, you use functions on them. R comes with all sorts of functions that you can apply to your counts data.
```{r}
sum(counts)
length(counts)
sum(counts)/length(counts)
```

For this example, the `sum()` function returns the total number of hurricanes making landfall. The `length()` function returns the number of years, and `sum(counts)/length(counts)` returns the average number of hurricanes per year.

Other useful functions include, `sort()`, `min()`, `max()`, `range()`, `diff()`, and `cumsum()`. Try these on the landfall counts.  What does `range()` do?  What does `diff()` do?

## Mean

The average (or mean) value of a set of numbers ($x$'s) is defined as:
$$
\bar x = \frac{x_1 + x_2 + \cdots + x_n}{n}
$$
The function `mean()` makes this calculation on your set of counts.
```{r}
mean(counts)
```

## Data vectors

The count data is stored as a vector. R keeps track of the order that the data were entered. First element,second element, and so on. This is good for a couple of reasons. Here the data has a natural order - year 1, year 2, etc. You don't want to mix these. You would like to be able to make changes to the data item by item instead of entering the entire data again. Also vectors are math objects making them easy to manipulate.

Suppose `counts` contain the annual number of land-falling hurricanes from the first decade of a longer record. You want to keep track of counts over other decades. This could be done by the following, example.
```{r}
cD1 <- counts
cD2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1) 
```
 
Note that you make a copy of the first decade of counts and save the vector using a different object name.

Most functions operate on each element of the data vector at the same time.
```{r}
cD1 + cD2
```

The first year of the first decade is added from the first year of the second decade and so on.

What happens if you apply the `c()` function to these two vectors?
```{r}
c(cD1, cD2)
```

If you are interested in each year's count as a difference from the decade mean, you type:
```{r}
cD1 - mean(cD1)
```

In this case a single number (the mean of the first decade) is subtracted from each element of the vector of counts.

This is an example of data recycling. R repeats values from one vector so that its length matches the other vector. Here the mean is repeated 10 times.

## Variance

Suppose you are interested in the variance of the set of landfall counts. The formula is given by:
$$
\hbox{var}(x) = \frac{(x_1 - \bar x)^2 + (x_2 - \bar x)^2 + \cdots + (x_n - \bar x)^2}{n-1}
$$

Note: The formula is given as LaTeX math code with the double dollar signs starting (and ending) the math mode. It's a bit hard to read but it translates exactly to math as you would read it in a scientific article or textbook. Look at the HTML file.

Although the `var()` function will compute this for you, here you see how you could do this directly using the vectorization of functions. The key is to find the squared differences and then add up the values.

The key is to find the squared differences and then add them up.
```{r}
x <- cD1
xbar <- mean(x)
x - xbar
(x - xbar)^2
sum((x - xbar)^2)
n <- length(x)
n
sum((x - xbar)^2)/(n - 1)
```

To verify type
```{r}
var(x)
```

## Data vectors have a type

One restriction on data vectors is that all the values have the same type. This can be numeric, as in counts, character strings, as in
```{r}
simpsons <- c("Homer", "Marge", "Bart", "Lisa", "Maggie")
simpsons
```

Note that character strings are made with matching quotes, either double, ", or single, '.

If you mix the type within a data vector, the data will be coerced into a common type, which is usually a character.  Arithmetic operations do not work on characters.

Returning to the land falling hurricane counts.
```{r}
cD1 <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)   
cD2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1)
```

Now suppose the National Hurricane Center (NHC) reanalyzes a storm, and that the 6th year of the 2nd decade is a 1 rather than a 0 for the number of landfalls. In this case you type
```{r}
cD2[6] <- 1    # assign the 6 year of the decade a value of 1 landfall
```

The assignment to the 6th entry in the vector cD2 is done by referencing the 6th entry of the vector with square brackets `[]`. 

It's important to keep this in mind: Parentheses `()` are used for functions and square brackets `[]` are used to extract values from vectors (and arrays, lists, etc). REPEAT: `[]` are used to extract or subset values from vectors, data frames, matrices, etc.

```{r}
cD2    #print out the values
cD2[2]  # print the number of landfalls during year 2 of the second decade
cD2[4]  # 4th year count
cD2[-4]  # all but the 4th year
cD2[c(1, 3, 5, 7, 9)]   # print the counts from the odd years
```

One way to remember how to use functions is to think of them as pets.  They don't come unless they are called by name (spelled properly). They have a mouth (parentheses) that likes to be fed (arguments), and they will complain if they are not feed properly.

## Working smarter

R's console keeps a history of your commands. The previous commands are accessed using the up and down arrow keys. Repeatedly pushing the up arrow will scroll backward through the history so you can reuse previous commands.

Many times you wish to change only a small part of a previous command, such as when a typo is made.  With the arrow keys you can access the previous command then edit it as desired.